Assessing the Accessibility of Protein Tunnels using Motion Planning

### 2017-12-09 paper.tex
corections

Reactivity of proteins with other molecules plays a crucial role in many research disciplines, including protein engineering and drug design.
Protein reactivity is tightly connected with the presence of the void space between its amino acids. 
This space, forming a path called tunnel, can be utilized by a small molecule (ligand) which can enter the protein inner space and react with its amino acids surrounding a specific site, called the active site~\cite{gora2013gates}.
\textcolor{blue}{The size and location of tunnels determine the ligand size and shape which can enter the active site.}
Currently the biochemists are using various computational methods for tunnel detection (Fig.~\ref{fig::motiv}) in order to estimate the possibility of interactions between a protein and a ligand.
This helps them to reduce the number of necessary in-vitro lab experiments.
The most common tools detect the tunnels using Voronoi diagrams~\cite{yaffe2008,caver3}, assuming a spherical ligand (probe), 
\textcolor{blue}{i.e., they approximate the ligand by a bounding sphere.}
However, the ligands are typically of a non-spherical shape which is completely ignored by these methods.

@@

Our approach aims to overcome this problem by taking into account the actual shape of the ligand.
This necessary requires to consider also conformational changes.
%Our approach aims to overcome this problem by taking into account the actual shape of the ligand and also its conformations.
%\textcolor{blue}{To provide the biochemists with a solution which detects tunnels also in these frames and understand the ligand passage, it is necessary to take into account not only the ligand shape but also its conformation changes when the ligand moves.}
%The proposed solution needs also to cope with the narrow passage problem, as proteins are dense structures limiting the ligand movements inside.
Such a ligand, changing the mutual position of its atoms, is denoted as a flexible ligand.

@@

Finding trajectories only along a given tunnel tunnel is faster than computing any exit-pathway from the protein due to decreased
volume of the configuration space to be searched.
Another advantage of our method in comparison with the existing approaches is that we enable the users to trace the ligand path leading from the
\textcolor{blue}{outside/outer?} solvent towards the protein active site.

@@

Most of the available tools for tunnel detection are based on Voronoi diagrams~\cite{yaffe2008,caver3} or they utilize a grid-based computation~\cite{sehnal2013mole}.
%The tunnels are pathways computed for a single atom (spherical probe) using Voronio-diagrams~\cite{yaffe2008,caver3} or 
%grid-based approaches~\cite{sehnal2013mole,petrek2006caver}.
The tunnels are represented as a sequence of spheres and characterized by their bottleneck, length, curvature, and a list of surrounding residues, whose physico-chemical properties are crucial for assessing the interaction possibilities between the protein and ligand.
More details can be found in the recently published state-of-the-art report about analysis and visualization of biomolecular cavities~\cite{Krone_2016}.
The main disadvantage of both grid-based and VD-based methods is that the shape of the ligand is not taken into account during the tunnel detection and it is therefore not easy to estimate if (and how) a non-spherical ligand might traverse the tunnel.
The ligand flexibility cannot be principally considered by these approaches.
Simulations based on molecular dynamics (MD) can directly evaluate the migration of a ligand into a protein, but they are significantly more computationally demanding~\cite{kingsley2014including}. 
Moreover, MD simulations have to be prepared for a specific ligand/protein pair, which is not practical for testing a large number of ligands.
The geometric-based methods are therefore still preferable due to their speed.

The traversability of a tunnel by a flexible ligand can be formulated as a motion planning problem and solved using sampling-based motion planners.
These planners can cope with objects (robots) with many Degrees of Freedom (DOF)  and they can also consider arbitrary geometry.
Two widely used planners are Probabilistic Roadmaps (PRM)~\cite{kavrakiForPP} and Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT}.
%Sampling-based planners like RRT can be used to find the trajectories even for flexible ligands, as they can cope with many-DOF systems and they can consider arbitrary shape.
%The computation of the ligand trajectories can be solved also using motion 
%In order to determine if the ligand can pass the tunnel using the geometric-based methods, it is necessary to compute a trajectory considering the shape of the ligand and its 3D translation and rotation.
%In each iteration of RRT, a random configuration $\qrand \in \C$ is generated and its nearest node $\qnear \in \T$ in the tree is found.
%A new configuration $\qnew$ is constructed on the line connecting $\qnear$ and $\qrand$ in the distance $\varepsilon$ from $\qnear$.
%If $\qnew$ is collision-free, it is added to the tree.
%The algorithm terminates if the tree approaches the goal configuration close enough.
%rrt for molecular: \cite{al2012motion}
%survey \cite{gipson2012computational}
%Solutions for tunnel detection using sampling-based methods however have not been discussed yet.
%Sampling-based planners have been used in various applications in robotics~\cite{elbanhawi2014sampling} and also  %,latombe1999motion} and also
% to study proteins, e.g. in  
%loop motions~\cite{cortes2004geometric},
%protein folding~\cite{amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro},
%protein folding~\cite{raveh2009rapid,novinskaya2015improving},
%or protein folding combined with ligand diffusion~\cite{cortes2010simulating}.
The well known issue of the sampling-based planners is the narrow passage problem.
Narrow passage is a small region in the configuration space, whose removal changes the connectivity of the space.
Due to its low volume, the probability of sampling the narrow passage is low.
Consequently, many iterations are needed in order to put enough samples into the passages, which increases the computation time.
PRM-based planners can cope with the narrow passage problem by increasing the probability of sampling in difficult regions, e.g.
by sampling along medial axis~\cite{amatoOBRRT,amato2002using,wilmarthMAPRM}.
In~\cite{bergWIG}, workspace is represented by a grid in which difficult regions are identified using watershed labeling.
The regions can be estimated online, e.g., based on the number of collision-free samples around a given configuration~\cite{overmarsGauss,hsuBridge} or even identified by a human operator~\cite{denny2018general}. 
%The narrow passages are however located in the configuration space and it is generally not easy to estimate their location
%based only on the features of the workspace~\cite{hannaWIS}.
%.g. based on amount of collision-free samples around a given sample~\cite{overmarsGauss} or e.g. based
%~\cite{hsuBridge,overmarsGauss}.

@@

The probability of sampling of the narrow passages can be increased by dilating the free-space, e.g., by shrinking the geometry of 
the robot~\cite{hsuOnProb} or the obstacles~\cite{bayazitIRC}.
The shrinking technique has been used also in~\cite{cortes2010simulating}, where the exit pathways for a small flexible molecule are
computed using a RRT-based method.
The flexible ligand is modeled as a kinematic chain, which increases the dimension of the configuration space.
To cope with these many-DOF ligands, the RRT-ML~\cite{cortes2007mlrrt} approach is employed as the basic planner in~\cite{cortes2010simulating}.
RRT-ML expands the tree primarily using those DOF that are essential for achieving the motion of the ligand (i.e., rotation
and translation) and it employs the other DOFs (i.e., those that are responsible for \textcolor{blue}{conformation/al?} changes) if they hinder the growth of the tree.
In~\cite{singhLIG}, the pathways for a flexible ligand are found using PRM.
Each configuration in the roadmap is also evaluated using electrostatic energy.
The high-energy configurations are discarded, while low-energy conformations (and their vicinity) are sampled more dense.
The approach~\cite{singhLIG} can be also used to predict ligand binding sites.

@@

In contrast to~\cite{cortes2010simulating}, which searches for any exit path from the protein, we localize the searching and compute the trajectories only along each tunnel separately which allows biochemists to analyze selected tunnels.
%This is motivated by the practical needs of biochemists that are interested in the actual traversability the ligand through a selected tunnel.
Computing the trajectories along a single tunnel brings several advantages from the motion planning point of view as well.
By selecting a tunnel to be analyzed, the users help the planner to focus only on a subset of the whole configuration
space, similarly as in~\cite{denny2018general}.
Consequently, the volume of the configuration space to be searched is decreased, while the relative volume of the narrow passages is increased.
The random samples are not generated in the whole configuration space, but only around the tunnel in a user-defined distance which helps to cope with the narrow passage problem.
%The tunnels serve as a navigation path for the configuration tree~\cite{vonasek2009rrt}.

Due to the protein dynamics, the tunnels move and merge, and their bottleneck can change in each time step.
Therefore, it can be expected that the detected tunnels are rather narrow.
Based on the observations of MD simulations, tunnels adapt their shape to the passing ligands and vice versa.
In consequence, even very narrow tunnels, being too small for the passage of a given ligand-approximating sphere, can serve as the transportation path for this ligand.
Therefore, also the existing Voronoi-based methods enable the user to decrease the size of the approximating sphere which simulates shrinking of the ligand size.
In our solution we simulate this by shrinking the atom radii of the ligand, used also by~\cite{cortes2010simulating,guieysse2008structure}.

@@

Proteins and ligands are represented by the hard sphere model, where the radius of each sphere representing an atom is given by its van der Waals radius.
A protein tunnel is described by a sequence of collision-free spheres 
$T=( (c_1, r_1),\ldots,(c_n,r_n) )$, where $n$ denotes the number of spheres,
$c_i \in \R^3$ is their 3D position, and $r_i > 0$ denotes the maximum collision-free radius of a sphere centered at $c_i$. 
Tunnels serving as an input for our proposed method can be detected using some of the existing computational tools, such as CAVER 3.0~\cite{caver3}.
%Conformations of the ligand may be generated in the absence of the receptor and subsequently docked[13] 
%or conformations may be generated on-the-fly in the presence of the receptor binding cavity,[14] 
%or with full rotational flexibility of every dihedral angle using fragment based docking.[15] 
%Force field energy evaluation are most often used to select energetically reasonable conformations,[16] but 
%knowledge-based methods have also been used.[17]
%13 = https://link.springer.com/article/10.1007/BF00123666
%The proteins are dense structures and the tunnels are typically narrow.
%Depending on the protein, the tunnel bottlenecks can be even less than $1$~\AA, which is too narrow for ligands with more than 2 atoms.
The ligand is modeled using the set $\L$ of its conformations.
Candidate conformations are prepared by discrete sampling of degrees of freedom of the ligand.
Each candidate conformation is ranked according its energy and only low-energy conformation are used in $\L$.
%The conformations are prepared using the Rosetta tool and ranked according their potential energy. Only conformations
%with energy lower than a user-defined threshold are used.
%The flexibility of the ligand is modeled using the set $\L$ of its conformations.
%We assume that the library of all possible conformations in a considered resolution is available
%(e.g., ~\cite{dunbrack}) 
%and we later show how to select a subset of conformations suitable for the motion planning.
To enable the motion of ligands in the narrow tunnels, the atomic radii of the ligand are scaled down by a factor $s, 0 < s \le 1$.
In our solution we use a discrete set of scales, i.e., $s \in \S=\{\smin, \smin+\sdelta, \ldots, \smax\}$, where 
$\smin$ is the minimal allowed scale, $\smax=1$ is the maximal allowed scale, and $\sdelta$ is the minimal difference between two scales.

@@

\linesnumbered
\begin{algorithm}[bh]
{\small
\setstretch{0.88}
\caption{\label{alg::main}Main loop of the RRT planner}
\KwIn{
    tunnel $T=( (c_i, r_i) )$, $i=1,\ldots,n$, with spheres centers $c_i \in \R^3$ and radii $r_i$,
    initial configuration $\qinit$
}
\KwData{
   ligand conformations $\L$,
   scale limits $\smin, \smax$ and $\sdelta$,
   distance $\dt$ to move virtual goal
}
\KwOut{
    configuration tree $\T$\;
}
\hrule
$v = 1$; // index of the virtual goal\\
$iteration = 0$\;
$\T$.addNode($\qinit$)\;
\While{$iteration < \Imax$ {\bf and}  $v < n$}{
    \eIf{$rand() < \gb$}{
        $\qrand$ = random sample around virtual goal $c_v\!\in\!T$\;
    }{
        $\qrand$ = random sample from $\C$\;
    }
    $\qnear$ = nearest node in $\T$ towards $\qrand$\;
    expand($\T, \qnear,\qrand$)\;
    \For{$i= n-1,n-2,\ldots,v+1,v$}{ \nllabel{alg::main:a}
        $d$ = nearest node in the tree towards sphere $c_i$\;
        \If{$d < \dt $}{
            $v = i+1$; // new virtual goal found\;
            {\bf  break}\;
        }
    } \nllabel{alg::main:b}
    $iteration = iteration+1$\;
}
\return $\T$\;
}
\end{algorithm}

@@

The atom-based metric can be evaluated fast using KD-trees with complexity $\mathcal{O}(\log n)$, where $n$ is the number of atoms of the ligand.
For each conformation $\l \in \L$ of the ligand, a KD-tree is built for the ligand at the configuration $q=(0,0,0,0,0,0,l,\smax)$
(i.e., for ligand with no translation and no rotation).
To find the nearest atom to $\qrand$ for a ligand at the configuration $q=(x,y,z,r_x,r_y,r_z,l,s)$, the translation vector $V=(x,y,z)$ and
rotation matrix $R$ given by $r_x$, $r_y$ and $r_z$ are computed.
Then, $\qrand$ is translated to $\qrand' = R^{-1}(\qrand-V)$ and the KD-tree search is used to find the nearest point towards $\qrand'$, which
is also the nearest point of the ligand at $q$ to $\qrand$.

@@

\begin{table}[ht]
\centering
\caption{\label{tab::selection}
    \small
    The influence of the conformation selection method on the traversability rates in the first (main) tunnel.
    The number after~`$/$' denotes the number of atoms.
    The protein has 4650 atoms.
}
\small
\renewcommand{\tabcolsep}{4.3pt}
{\scriptsize
\renewcommand{\arraystretch}{0.7}
\input graphSelNew.tex
}
\end{table}

@@

\begin{table}[bt]
\centering
\caption{\label{tab::main}
    \small
    Traversability rate [\%] over 100 frames for ligands with 50 conformations. 
    The number after '$/$' denotes the number of atoms.
    The protein has 4650 atoms.
}
\small
\renewcommand{\tabcolsep}{2pt}
{
\scriptsize
\renewcommand{\arraystretch}{0.7}
\input graphNew.tex
}
\end{table}

The traversability rate is shown in Tab.~\ref{tab::main}. 
The table shows the traversability rate only for the main tunnel, as no trajectory was found in the side tunnel for most of the tested ligands.
The only exception is the ligand m037t, which was the smallest tested one (with 8 atoms), and therefore it could fit even inside the side tunnel.
The traversability rate is higher for the most scaled-down ligands ($\smin=0.5$) and it decreases with the increasing scale.
For the ligand DCP, trajectories were detected in 18~\% of frames for scale $\smin=0.8$ and the traversability of this ligand
to the active site was approved by a separate MD simulation~\cite{marques2017catalytic}.
%For the scale $\smin=0.8$, the most traversable frames were detected for ligands DCP (18~\%), m037t (45~\%) and m038t (30~\%).
%The traversability of the DCP ligand to active site was approved by a separate MD simulation~\cite{marques2017catalytic}.
Based on the number of success rate, also other ligands (e.g., m037t with traversability rate 45~\% at $\smin$=0.8) seem to be promising.
However, their binding was not approved yet by MD dedicated simulations.

>>>>

Reactivity of proteins with other molecules plays a crucial role in many research disciplines, including protein engineering and drug design.
Protein reactivity is tightly connected with the presence of the void space between its amino acids. 
This space, forming a path called tunnel, can be utilized by a small molecule (ligand) which can enter the protein inner space and react with its amino acids surrounding a specific site, called the active site~\cite{gora2013gates}.
\textcolor{blue}{The size and location of tunnels determine the ligand size and shape which can enter the active site.}
Currently, the biochemists use various computational methods for tunnel detection (Fig.~\ref{fig::motiv}) in order to estimate the possibility of interactions between a protein and a ligand.
This helps them to reduce the number of necessary in-vitro lab experiments.
The most common tools detect the tunnels using Voronoi diagrams~\cite{yaffe2008,caver3} assuming a spherical ligand (probe), 
i.e., they approximate the ligand by a bounding sphere.
However, the ligands are typically of a non-spherical shape which is completely ignored by these methods.

@@

Our approach aims to overcome this problem by taking into account the actual shape of the ligand.
This necessarily requires to consider also conformational changes, i.e., the mutual position of the ligand atoms can change.
Such a ligand is denoted as flexible ligand.
%Our approach aims to overcome this problem by taking into account the actual shape of the ligand and also its conformations.
%\textcolor{blue}{To provide the biochemists with a solution which detects tunnels also in these frames and understand the ligand passage, it is necessary to take into account not only the ligand shape but also its conformation changes when the ligand moves.}
%The proposed solution needs also to cope with the narrow passage problem, as proteins are dense structures limiting the ligand movements inside.
%Such a ligand, for which the mutual position of its atoms can change, is denoted as a flexible ligand.

@@

Finding trajectories only along a given tunnel tunnel is faster than computing any exit pathway from the protein due to decreased
volume of the configuration space to be searched.
Another advantage of our method in comparison with the existing approaches is that we enable the users to trace the ligand path leading from the
\textcolor{blue}{outside/outer?} solvent towards the protein active site.

@@

Most of the available tools for tunnel detection are based on Voronoi diagrams~\cite{yaffe2008,caver3} or they utilize a grid-based computation~\cite{sehnal2013mole}.
%The tunnels are pathways computed for a single atom (spherical probe) using Voronio-diagrams~\cite{yaffe2008,caver3} or 
%grid-based approaches~\cite{sehnal2013mole,petrek2006caver}.
The tunnels are represented as a sequence of spheres and characterized by their bottleneck, length, curvature, and a list of surrounding residues, whose physico-chemical properties are crucial for assessing the interaction possibilities between the protein and ligand.
More details can be found in the recently published state-of-the-art report about analysis and visualization of biomolecular cavities~\cite{Krone_2016}.
The main disadvantage of both grid-based and VD-based methods is that the shape of the ligand is not taken into account during the tunnel detection and it is therefore not easy to estimate if (and how) a non-spherical ligand can traverse the tunnel.
The ligand flexibility cannot be principally considered by these approaches.
Simulations based on molecular dynamics (MD) can directly evaluate the migration of a ligand into a protein, but they are significantly more computationally demanding~\cite{kingsley2014including}. 
Moreover, MD simulations have to be prepared for a specific ligand/protein pair, which is not practical for testing a large number of ligands.
The geometric-based methods are therefore still preferable due to their speed.

The traversability of a tunnel by a flexible ligand can be formulated as a motion planning problem and solved using sampling-based motion planners.
These planners can cope with objects (robots) with many Degrees of Freedom (DOF)  and they can also consider objects of arbitrary shape.
Two widely used planners are Probabilistic Roadmaps (PRM)~\cite{kavrakiForPP} and Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT}.
%Sampling-based planners like RRT can be used to find the trajectories even for flexible ligands, as they can cope with many-DOF systems and they can consider arbitrary shape.
%The computation of the ligand trajectories can be solved also using motion 
%In order to determine if the ligand can pass the tunnel using the geometric-based methods, it is necessary to compute a trajectory considering the shape of the ligand and its 3D translation and rotation.
%In each iteration of RRT, a random configuration $\qrand \in \C$ is generated and its nearest node $\qnear \in \T$ in the tree is found.
%A new configuration $\qnew$ is constructed on the line connecting $\qnear$ and $\qrand$ in the distance $\varepsilon$ from $\qnear$.
%If $\qnew$ is collision-free, it is added to the tree.
%The algorithm terminates if the tree approaches the goal configuration close enough.
%rrt for molecular: \cite{al2012motion}
%survey \cite{gipson2012computational}
%Solutions for tunnel detection using sampling-based methods however have not been discussed yet.
%Sampling-based planners have been used in various applications in robotics~\cite{elbanhawi2014sampling} and also  %,latombe1999motion} and also
% to study proteins, e.g. in  
%loop motions~\cite{cortes2004geometric},
%protein folding~\cite{amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro},
%protein folding~\cite{raveh2009rapid,novinskaya2015improving},
%or protein folding combined with ligand diffusion~\cite{cortes2010simulating}.
The well known issue of the sampling-based planners is the narrow passage problem.
Narrow passage is a small region in the configuration space, whose removal changes the connectivity of the space.
Due to its low volume, the probability of sampling the narrow passage is low.
Consequently, many iterations are needed in order to put enough samples into the passages, which increases the computation time.
PRM-based planners can cope with the narrow passage problem by increasing the probability of sampling in difficult regions, e.g.
by sampling along medial axis~\cite{amatoOBRRT,amato2002using,wilmarthMAPRM}.
In~\cite{bergWIG}, workspace is represented by a grid in which difficult regions are identified using watershed labeling.
The regions can be estimated online, e.g., based on the number of collision-free samples around a given configuration~\cite{overmarsGauss,hsuBridge} or even identified by a human operator~\cite{denny2018general}. 
%The narrow passages are however located in the configuration space and it is generally not easy to estimate their location
%based only on the features of the workspace~\cite{hannaWIS}.
%.g. based on amount of collision-free samples around a given sample~\cite{overmarsGauss} or e.g. based
%~\cite{hsuBridge,overmarsGauss}.

@@

The probability of sampling of the narrow passages can be increased by dilating the free-space, e.g., by shrinking the geometry of 
the robot~\cite{hsuOnProb} or the obstacles~\cite{bayazitIRC}.
The shrinking technique has been used also in~\cite{cortes2010simulating}, where the exit pathways for a small flexible molecule are
computed using a RRT-based method.
The flexible ligand is modeled as a kinematic chain, which increases the dimension of the configuration space.
To cope with these many-DOF ligands, the RRT-ML~\cite{cortes2007mlrrt} approach is employed as the basic planner in~\cite{cortes2010simulating}.
RRT-ML expands the tree primarily using those DOF that are essential for achieving the motion of the ligand (i.e., rotation
and translation) and it employs the other DOFs (i.e., those that are responsible for conformational changes) if they hinder the growth of the tree.
In~\cite{singhLIG}, the pathways for a flexible ligand are found using PRM.
Each configuration in the roadmap is also evaluated using electrostatic energy.
The high-energy configurations are discarded, while low-energy conformations (and their vicinity) are sampled more dense.
The approach~\cite{singhLIG} can be also used to predict ligand binding sites.

@@

In contrast to~\cite{cortes2010simulating}, which searches for any exit pathway from the protein, we localize the searching and compute the trajectories only along each tunnel separately which allows biochemists to analyze selected tunnels.
%This is motivated by the practical needs of biochemists that are interested in the actual traversability the ligand through a selected tunnel.
Computing the trajectories along a single tunnel brings several advantages from the motion planning point of view as well.
The random samples are not generated in the whole configuration space, but only around the tunnel in a user-defined distance which helps to cope with the narrow passage problem.
%The tunnels serve as a navigation path for the configuration tree~\cite{vonasek2009rrt}.
By selecting a tunnel to be analyzed, the users help the planner to focus only on a subset of the whole configuration
space, similarly as in~\cite{denny2018general}.
Consequently, the volume of the configuration space to be searched is decreased, while the relative volume of the narrow passages is increased.

Due to the protein dynamics, the tunnels move and merge, and their bottleneck can change in each time step.
It can be expected that the tunnels in MD simulations without ligands are rather narrow.
However, the tunnels adapt their shape to the passing ligands and vice versa.
%Therefore, it can be expected that the detected tunnels are rather narrow.
In consequence, even very narrow tunnels, being too small for the passage of a given ligand-approximating sphere, can serve as the transportation path for the ligand.
Therefore, also the existing Voronoi-based methods enable the user to decrease the size of the approximating sphere which simulates shrinking of the ligand size.
In our solution we simulate this by shrinking the atom radii of the ligand, used also by~\cite{cortes2010simulating,guieysse2008structure}.

@@

Proteins and ligands are represented by the hard sphere model, where the radius of each sphere representing an atom is given by its van der Waals radius.
A protein tunnel is described by a sequence of collision-free spheres 
$T=( (c_1, r_1),\ldots,(c_n,r_n) )$, where $n$ denotes the number of spheres,
$c_i \in \R^3$ is their 3D position, and $r_i > 0$ denotes the maximum collision-free radius of a sphere centered at $c_i$. 
Tunnels serving as an input for the proposed method can be detected using some of the existing computational tools, such as CAVER 3.0~\cite{caver3}.
%Conformations of the ligand may be generated in the absence of the receptor and subsequently docked[13] 
%or conformations may be generated on-the-fly in the presence of the receptor binding cavity,[14] 
%or with full rotational flexibility of every dihedral angle using fragment based docking.[15] 
%Force field energy evaluation are most often used to select energetically reasonable conformations,[16] but 
%knowledge-based methods have also been used.[17]
%13 = https://link.springer.com/article/10.1007/BF00123666
%The proteins are dense structures and the tunnels are typically narrow.
%Depending on the protein, the tunnel bottlenecks can be even less than $1$~\AA, which is too narrow for ligands with more than 2 atoms.
The ligand is modeled using the set $\L$ of its conformations.
Candidate conformations are prepared by discrete sampling of degrees of freedom of the ligand.
Each candidate conformation is ranked according its energy and only low-energy conformations are used in $\L$.
%The conformations are prepared using the Rosetta tool and ranked according their potential energy. Only conformations
%with energy lower than a user-defined threshold are used.
%The flexibility of the ligand is modeled using the set $\L$ of its conformations.
%We assume that the library of all possible conformations in a considered resolution is available
%(e.g., ~\cite{dunbrack}) 
%and we later show how to select a subset of conformations suitable for the motion planning.
To enable the motion of ligands in the narrow tunnels, the atomic radii of the ligand are scaled down by a factor $s, 0 < s \le 1$.
In our solution we use a discrete set of scales, i.e., $s \in \S=\{\smin, \smin+\sdelta, \ldots, \smax\}$, where 
$\smin$ is the minimal allowed scale, $\smax=1$ is the maximal allowed scale, and $\sdelta$ is the minimal difference between two scales.

@@

\linesnumbered
\begin{algorithm}[hb]
{\small
\setstretch{0.88}
\caption{\label{alg::main}Main loop of the RRT planner}
\KwIn{
    tunnel $T=( (c_i, r_i) )$, $i=1,\ldots,n$, with spheres centers $c_i \in \R^3$ and radii $r_i$,
    initial configuration $\qinit$
}
\KwData{
   ligand conformations $\L$,
   scale limits $\smin, \smax$ and $\sdelta$,
   distance $\dt$ to move virtual goal
}
\KwOut{
    configuration tree $\T$\;
}
\hrule
$v = 1$; // index of the virtual goal\\
$iteration = 0$\;
$\T$.addNode($\qinit$)\;
\While{$iteration < \Imax$ {\bf and}  $v < n$}{
    \eIf{$rand() < \gb$}{
        $\qrand$ = random sample around virtual goal $c_v\!\in\!T$\;
    }{
        $\qrand$ = random sample from $\C$\;
    }
    $\qnear$ = nearest node in $\T$ towards $\qrand$\;
    expand($\T, \qnear,\qrand$)\;
    \For{$i= n-1,n-2,\ldots,v+1,v$}{ \nllabel{alg::main:a}
        $d$ = nearest node in the tree towards sphere $c_i$\;
        \If{$d < \dt $}{
            $v = i+1$; // new virtual goal found\;
            {\bf  break}\;
        }
    } \nllabel{alg::main:b}
    $iteration = iteration+1$\;
}
\return $\T$\;
}
\end{algorithm}

@@

The atom-based metric can be evaluated fast using KD-trees with complexity $\mathcal{O}(\log n)$, where $n$ is the number of atoms of the ligand.
For each conformation $\l \in \L$ of the ligand, a KD-tree is built for the ligand at the configuration $q=(0,0,0,0,0,0,l,\smax)$
(i.e., for ligand with no translation and no rotation).
To find the nearest atom to $\qrand$ for a ligand at the configuration $q=(x,y,z,r_x,r_y,r_z,l,s)$, the translation vector $V=(x,y,z)$ and
rotation matrix $R$ given by $r_x$, $r_y$ and $r_z$ are computed.
Then, $\qrand$ is translated to $\qrand' = R^{-1}(\qrand-V)$ and the KD-tree search is used to find the nearest atom from $q$ towards atoms of $\qrand'$.
%is also the nearest point of the ligand at $q$ to $\qrand$.

@@

\begin{table}[ht]
\centering
\caption{\label{tab::selection}
    \small
    The influence of the conformation selection method on the traversability rates in the first (main) tunnel.
    The number after~`$/$' denotes the number of atoms.
    The protein has 4650 atoms. 
    The rate is shown in [\%].
}
\small
\renewcommand{\tabcolsep}{4.3pt}
{\scriptsize
\renewcommand{\arraystretch}{0.7}
\input graphSelNew.tex
}
\end{table}

@@

\begin{table}[bt]
\centering
\caption{\label{tab::main}
    \small
    Traversability rate [\%] over 100 frames for ligands with 50 conformations. 
    The number after '$/$' denotes the number of atoms.
    The protein has 4650 atoms.
    The rate is shown in [\%].
}
\small
\renewcommand{\tabcolsep}{2pt}
{
\scriptsize
\renewcommand{\arraystretch}{0.7}
\input graphNew.tex
}
\end{table}

The traversability rate is shown in Tab.~\ref{tab::main}. 
The table shows the traversability rate only for the main tunnel, as no trajectory was found in the side tunnel for most of the tested ligands.
The only exception is the ligand m037t, which was the smallest tested one (with 8 atoms), and therefore it could fit even inside the side tunnel.
The traversability rate is higher for the most scaled-down ligands ($\smin=0.5$) and it decreases with the increasing scale.
For the ligand DCP, trajectories were detected in 18~\% of frames for scale $\smin=0.8$ and the traversability of this ligand
to the active site was approved by a separate MD simulation~\cite{marques2017catalytic}.
%For the scale $\smin=0.8$, the most traversable frames were detected for ligands DCP (18~\%), m037t (45~\%) and m038t (30~\%).
%The traversability of the DCP ligand to active site was approved by a separate MD simulation~\cite{marques2017catalytic}.
Based on the number of success rate, also other ligands (e.g., m037t with traversability rate 45~\% at $\smin$=0.8) seem to be promising.
However, their binding was not approved yet by dedicated MD simulations.
